Using Scagnostics to find Interesting Projections of Multivariate Data

### 2018-09-26 scagnostics.Rmd
Added first neutron star example.

## Data description
The model describes a neutron star merger and contains 6 free parameters, i.e. each neutron star is described by its mass and radius, and a so-called tidal deformability parameter $\Lambda$. A posterior sample containing 2538 points is available @ligoData and can be used to constrain the parameters. Here we use the "Parametrized-EoS\_nomaxmass\_posterior\_samples.dat" dataset. For details about the modelling and statistical procedure see @Abbott2018exr. A common representation of posterior samples used for interpretatin are so-called "corner plots", i.e. density displays for all 2-d parameter combinations as well as the 1-d profile, highlighting empirically evaluated regions of highest posterior density, which correspond to credibility regions at a given level.
Use Rorys script to make nice version of his corner plot.
Here we generate such a corner plot using the python package "ChainConsumer", highlighting the x and y % CIs. We point out the following features seen from Fig. XX: the combination of the two mass paremters is well constrained (a consequence of how a particular combination, the "chirp mass" enters predictions), other parameters also show patterns of dependence, they are however less pronounces. We highlight the non-linear relation between the radius $R_1$ and $\Lambda$ parameter and the two-pronged structure of the radius $R_1$.
For further study we therefore remove m2 from the dataset. We first look at the distribution in a grand tour display, where we use a scatter plot display combined with alpha transperancy to study the density distribution in the 5-d parameter space. The grand tour indicates that the data points are found on a curved surface in the five dimensional space. This may stem from a non-linear relation between the parameters, which we will try to uncover using the above developped formalism for new guided tour index functions.

## Using the guided tour
Given the observations from the grand tour display we expect a functional relation between some of the parameters, and we therefore optimise the "splines2d" index function. Note that all parameters are standardised to the range $[0,1]$ before running the optimisation.
Show final view found from the guided tour, give corresponding projection matrix, make corresponding plot from combination of original parameters.

>>>>

## Data description
The model describes a neutron star merger and contains 6 free parameters, i.e. each neutron star is described by its mass and radius, and a so-called tidal deformability parameter $\Lambda$. A posterior sample containing 2538 points is available @ligoData and can be used to constrain the parameters. Here we use the "Parametrized-EoS\_nomaxmass\_posterior\_samples.dat" dataset. For details about the modelling and statistical procedure see @Abbott2018exr. A common representation of posterior samples used for interpretatin are so-called "corner plots", i.e. density displays for all 2-d parameter combinations as well as the 1-d profile, highlighting empirically evaluated regions of highest posterior density, which correspond to credibility regions at a given level. The information can similarly be presented in a scatter plot matrix using transparency to convey the density information. Since in the following we work with 2-d scatter plots we chooose to show the data in this fashion, see Fig.

```{r neutronStarSPLOM}
nsD <- read_csv("data/samples.csv") 
ggpairs(nsD, lower=list(continuous = wrap("points", alpha = 0.05)))
```

We point out the following features seen from Fig. XX: the combination of the two mass paremters is well constrained (a consequence of how a particular combination, the "chirp mass" enters predictions), other parameters also show patterns of dependence, they are however less pronounces. We highlight the non-linear relation between the radius $R_1$ and $\Lambda$ parameter and the two-pronged structure of the radius $R_1$.
For further study we therefore remove m2 from the dataset. We first look at the distribution in a grand tour display, where we use a scatter plot display combined with alpha transperancy to study the density distribution in the 5-d parameter space. The grand tour indicates that the data points are found on a curved surface in the five dimensional space. This may stem from a non-linear relation between the parameters, which we will try to uncover using the above developped formalism for new guided tour index functions.

## Using the guided tour
Given the observations from the grand tour display we expect a functional relation between some of the parameters, and we therefore optimise the "splines2d" index function. Note that all parameters are standardised to the range $[0,1]$ before running the optimisation.
Show final view found from the guided tour, give corresponding projection matrix, make corresponding plot from combination of original parameters.
```{r neutronStarEq,  results="hide"}
nsD <- read_csv("data/samples.csv") 
nsM <- nsD %>%
  select(-m2) %>% # remove m2 to avoid finding well known correlation
  rescale() # rescaling, this returns a matrix
nseTour <- save_history(nsM, guided_tour(splineIndex()))
nseTourFull <- as.list(interpolate(nseTour))
iLast <- length(nseTourFull)
fProj <- nseTourFull[[iLast]]
dProj <- as.tibble(nsM %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

```{r nseMatrix, echo=FALSE}
knitr::kable(fProj ,  caption = "Matrix representation of best projection.")
```

The resulting projection clearly shows how the data selects a region that is well defined as a combination of several parameters. From the matrix representation we read off that parameters m1, L1 and R1 are important in this projection. Using the original (non-rescaled) data and taking into account the difference in scale between m1 and R1 we can reproduce a similar picture shown in Fig \ref{nsePlotOrig} (right), which we compare to the next closest bivariate plot shown on the left. It is clear that in this example additonal information is found when considering general projections rather than focussing only on bivariate relations.

```{r nsePlotOrig}
p1 <- ggplot(nsD, aes(R1, L1)) + geom_point(alpha = 0.05)
p2 <- ggplot(nsD, aes(R1-2*pi*m1, L1)) + geom_point(alpha = 0.05)
grid.arrange(p1, p2, ncol=2)
```
